---
title: "Sort Data"
author: "Shu Guo"
date: "Thursday, February 26, 2015"
output: pdf_document
---

## Sort Data Using R
Here we will talk about sorting data using three R functions: `sort()`, `order`, and `arrange()` in `dplyr` package.

We can `sort()` function to sort a vector or factor into ascending or descending order.   
`sort(x, decreasing = FALSE, na.last = NA, ...)`  
We can set `decreasing = TRUE` to sort the object by descending order, `na.last` is used for controlling the treatment of NAs. IF TRUE, missing values in the data are put last; if FALSE, they are put first; if NA, they are removed.
```{r}
# sort() example
data(mtcars)
# sort a vector of numerical values
sort(mtcars$mpg)
# sort a factor variable
sort(factor(mtcars$cyl))
# sort the entire data set by descending order of mpg
sort(mtcars$mpg, decreasing = TRUE, na.last = TRUE)
```

So we can order a vector now, but what if we want to order a data set using one or more of the variables in it? Here comes the function `order()`:    
`order(..., na.last = TRUE, decreasing = FALSE)`

This function returns a vector of indices which rearranges its first argument into ascending or descending order.

```{r}
# order() example
order(mtcars$mpg)
# sort the data set by ascending values of mpg
head(mtcars[order(mtcars$mpg), ])
# sort the data set by ascending order of the factor variable cyl and descending order of mpg
head(mtcars[order(factor(mtcars$cyl), -mtcars$mpg), ], n = 15)

```